### 2024-03-14 README.Rmd
this?
further changes from checklist
improved readme

-   These do not have to be "literal" islands but any situation where discontiguous geographical units are present.

@@

@@

1.  The initial setting up neighbourhood structures can be frustrating for people who are eager to get started with fitting spatial models. This is especially so when the presence of discontiguities within a geographical dataset means that, even having set up a neighbours list, the model will still not run without further awkward data manipulations.

@@

3.  If there are some neighbours assigned which are not appropriate, or if you wish to add additional ones, there are functions to allow this to be done in a straightforward and openly reportable way.

4.  Once an appropriate neighbourhood structure is in place, different types of statistical tests and models can be performed. `sfdep` contains functionality to perform such test, and the output from `sfislands` can be used in its functions.

@@

The following is a framework within which the `sfislands` functions could be used.

@@

*Use the output of **st_bridges()** as both the data and neighbourhood inputs for a model using, for example, `mgcv`, `brms` or `inla`.*

@@

@@

## Examples

In the context of the constituencies of England, Scotland and Wales, `st_bridges` will define a neighbourhood structure which includes the islands. `st_quickmap_nb` will show a representation of this on a map:

```{r, echo=FALSE}
uk_election$country <- case_when(uk_election$region == "Wales" ~ "Wales",
                                  uk_election$region == "Scotland" ~ "Scotland",
                                  TRUE ~ "England")
```

```{r, fig.width=12, fig.height=10}
nbsf <- st_bridges(df = uk_election,
                   geom_col_name = "constituency_name",
                   link_islands_k = 2)
st_quickmap_nb(nbsf,
                    pointsize=0.05, 
                    title = "st_bridges() contiguities",
                    subtitle = "islands linked to k=2 nearest constituencies")

@@

Neighbours can be manually added or removed using `st_manual_join_nb()` and `st_manual_cut_nb()`.

```{r}
st_bridges(df = uk_election|> filter(region %in% c("Wales","South West")),
           geom_col_name = "constituency_name",
           link_islands_k = 2)  |> 
  st_manual_join_nb("Gower","St Ives") |> 
  st_quickmap_nb()
```

## Modelling & post-functions

Having set up a neighbourhood structure and embedded it as a named list or matrix within the original `sf` dataset as column `nb`, there are some functions to make it easy to quickly perform ICAR smoothing, augment the original dataframe with these predictions, and visualise them.

For example, we can use the `mgcv` package to generate a Markov Random Field ICAR smooth of poor health across the study area. This is done very quickly by using `st_bridges()` to prepare the data, putting that inside the `mgcv` GAM formulation, and then piping into the `st_augment()` and `st_quickmap_preds()` functions.

```{r, fig.width=12, fig.height=6}

prep_data3 <- st_bridges(uk_election, "constituency_name") # decide upon the contiguities and add them to the df

model <- gam(con_swing ~ 
               s(region, bs="re") + # region level random intercept
               s(county, bs="re") + # county level random intercept
               s(county, degree_educated, bs="re") + # county level random coefficient
               s(constituency_name, bs='mrf', 
                 xt=list(nb=prep_data3$nb),k=10) + # ICAR constituency ICAR varying coefficients
               s(constituency_name, by=white, bs='mrf', 
                 xt=list(nb=prep_data3$nb),k=10), # ICAR constituency ICAR varying coefficients
             data=prep_data3, method="REML") |> 
  st_augment(prep_data3) |> # pipe into function to get estimates
  st_quickmap_preds() # pipe into this for visualisation

ggarrange(plotlist = model, legend = "none", nrow=1)
```

Further information and worked examples are available in the vignettes which can be found [here](https://horankev.github.io/sfislands/).

>>>>

-   These do not have to be *islands* by a traditional *land and water* geographic definition, but any situation where discontiguous geographical units are present.

@@

Install the released version from CRAN

``` r
install.packages("sfislands")
```

@@

1.  The initial setting-up of neighbourhood structures can be frustrating for people who are eager to get started with fitting spatial models. This is especially so when the presence of discontiguities within a geographical dataset means that, even having set up a neighbours list, the model will still not run without further awkward data manipulations.

@@

3.  If some units have been assigned as neighbours inappropriately, or if we wish to add additional connections, there are functions to allow this to be done in a straightforward and openly reportable way.

4.  Once an appropriate neighbourhood structure is in place, different types of statistical tests and models can be performed. `sfdep` contains functionality to perform such tests, and the output from `sfislands` can be used in its functions.

@@

The following is a brief explanations of the functions contained in the `sfislands` package and a workflow within which they could be used.

@@

Use the output of `st_bridges()` as both the data and neighbourhood inputs for a model using, for example, `mgcv`, `brms` or `inla`.

@@

```
```{r, echo=FALSE}
uk_election$country <- case_when(uk_election$region == "Wales" ~ "Wales",
                                  uk_election$region == "Scotland" ~ "Scotland",
                                  TRUE ~ "England")
```

## Workflow

The dataset `uk_elections`, which is provided with the package, is used for the following demonstrations. It is an `sf` dataset with geometries for each of the 632 Westminster parliamentary constituencies of England, Scotland and Wales in the 2019 general election. For each constituency, it contains socio-economic data from the 2011 census. It also shows the *Butler swing* between the 2017 and 2019 elections, which is a measure of the change in voter allegiance in favour of the Conservative Party from the Labour Party.

### Pre-functions: Creating desired adjacency structure

To demonstrate the `st_bridges()` function for preparing an adjacency matrix, we focus on Scotland. It is the nation with the highest number of islands. The function, with default arguments, returns a dataframe with an additional column named `nb`. This column is a named neighbourhood list of class `nb`. The default neighbourhood structure which it computes is:

1.    for constituencies which are **not islands**, first-order queen contiguity. A constituency is considered to be a neighbour of all of those with which it shares at least a vortex.

2.    for **islands**, contiguities are assigned to the two nearest constituencies to each island. This number can be altered using the `link_islands_k` argument.

```{r}
nbsf <- st_bridges(df = uk_election |> filter(country == "Scotland"),
                   geom_col_name = "constituency_name",
                   link_islands_k = 2)
```

This new dataframe has the following form (with a column named `nb` containing a neighbours list):

```{r}
head(nbsf)
```
We can inspect (and openly report) what additional contiguities have been added to the standard queen contiguity structure using the function `st_check_islands()`.

```{r}
st_check_islands(nbsf)
```

The overall contiguity structure can be visualised using the function `st_quickmap_nb()`, whose output is shown below on the left.

```{r, fig.width=12, fig.height=10}
ggarrange(
  st_quickmap_nb(nbsf),

  st_quickmap_nb(nbsf,
                 concavehull = TRUE,
                 nodes = "numeric",
                 numericsize = 3.5),

  ncol = 2
)

@@

However, it is often the case that each *island* unit (in this case, constituency) is in fact a multipolygon of islands. To make this clearer, the argument `concavehull = TRUE` can be chosen. These green boundaries (shown above on the right) around each constituency are for reference purpose only and do not influence the assignment of contiguities. They show which group of *islands* are considered as one *constituency of islands*. Furthermore, the argument `nodes = "numeric"` represents the centroids of each unit with its index number instead of a geometric point. This can be useful for editing the adjacencies, particularly if the researcher is not familiar with the names of the individual spatial units.

#### Editing contiguities

A researcher can now use domain knowledge or a particular hypothesis to edit these contiguities. For instance, we may not be happy with the connection between 47 (Na h-Eileanan An Iar) and 51 (Orkney and Shetland) and wish to remove it. We might instead feel that it is important for 47 and 5 to be considered neighbours. This can be easily achieved, as shown below, using the `st_manual_cut_nb()` and `st_manual_join_nb()` functions. The spatial units whose relationships we wish to change can be referenced by name or by index number.

```{r}
nbsf <- st_bridges(df = uk_election |> filter(country == "Scotland"),
                   geom_col_name = "constituency_name",
                   link_islands_k = 2) |> 
  st_manual_cut_nb("Na h-Eileanan An Iar", "Orkney and Shetland") |> 
  st_manual_join_nb(47, 5)
```

The results of these changes as visualised in map form are shown below:

```{r, fig.width=12, fig.height=10}
ggarrange(
  st_quickmap_nb(nbsf,
                 concavehull = TRUE,
                 nodes = "numeric",
                 numericsize = 3.5),

  st_quickmap_nb(nbsf),

  ncol = 2
)

@@

The above process can be iterated until we are happy that the structure conforms with our hypothesis of connectivity within the study area.

### Modelling & post-functions

Having set up a neighbourhood structure and embedded it as a named list or matrix within the original `sf` dataset as a column named `nb`, the workflow to generate and visualise predictions from a model (in this case using `mgcv`) is very straightforward and consists of only 3-4 lines of code.

For example, we can use the `mgcv` package to generate quite a complicated model with various random intercepts and slopes at region and county level, and additionally, different Markov Random Field ICAR smooths at the constituency level. This is done very quickly by

* using `st_bridges()` to prepare the data, 
* putting that inside the `mgcv::gam()` formulation, and then 
* piping into the `st_augment()` function.

This gives the following output, with all of the spatially varying predictions and their standard errors being added as extra columns before the final geometry column:

```{r}
prep_data <- st_bridges(uk_election, "constituency_name") # decide upon the contiguities and add them to the df

model1 <- gam(con_swing ~ 
               s(region, bs="re") + # region level random intercept
               s(county, bs="re") + # county level random intercept
               s(county, degree_educated, bs="re") + # county level random coefficient
               s(constituency_name, bs='mrf', 
                 xt=list(nb=prep_data$nb),k=10) + # ICAR constituency ICAR varying coefficients
               s(constituency_name, by=white, bs='mrf', 
                 xt=list(nb=prep_data$nb),k=10), # ICAR constituency ICAR varying coefficients
             data=prep_data, method="REML") |>  # pipe into function to get estimates
  st_augment(prep_data)

head(model1)

The predictions can be quickly visualised using `st_quickmap_preds()`. As this function generates a list of `ggplots`, they can be conveniently visualised using the `ggarrange()` function from the `ggpubr` package.

```{r, fig.width=12, fig.height=6}
plots <- st_quickmap_preds(model1)
ggarrange(plotlist = plots, legend = "none", nrow=1)
```

To emphasise the simplicity of the process, the a modified version of the following sequence of commands will generate a series of plots of all of the spatially varying elements of a model defined in `mgcv`:

```{r, eval=FALSE}
prep_data <- st_bridges(...)

gam(..., data = prep_data) |> 
  st_augment() |> 
  st_quickmap_preds()

It is also possible to conduct the same kind of augmenting and plotting of models generated using the `lme4` and `nlme` packages.

Further information and vignettes containing more detailed examples are available [here](https://horankev.github.io/sfislands/).
